\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2015-12-07 paper/Paper_sweave.Rnw
Updated

\documentclass{article}

\begin{document}
\SweaveOpts{concordance=TRUE}

@@

\author{Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
and\\
Nathaniel Tomasetti\\
Department of Econometrics and Business Statistics, Monash University\\
  \maketitle
}

@@

\begin{multicols}{2}
\twocolumn

<<setup, echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)

@@

In order to work with data, it first must be understood. Statistical inference requires hypotheses to be established prior to data collection, but often data is collected first. This is especially so today, for vast databases that have been assembled in the big data era, that now need the data scientist to unravel the meaning of the numbers. Without preset hypotheses to test, the power of statistical inference is impotent, and without hypotheses the data analyst can stumble blindly trying to build up models of structure in the data.  To understand data requires good visualisation. This idea was formed as early as the 18th century, when William Playfair institutionalized the then revolutionary idea of graphing government and economic data. Far easier than reading tables of numbers, these ideas were powerful, and by providing the basic building blocks for plotting statistical data, his graphic designs became a conduit for the communication of otherwise complex information [20]. Since then, advances in computing power have allowed statistical graphics to flourish, spearheaded by Tukey in 1965 into the new domain of exploratory data analysis (EDA) [7, 23]. EDA can be thought of as a well understood [3, 25] set of tools and techniques required to visualise information, to physically see what the data contains. In particular, it incorporates a free roaming approach, where the analyst is able to explore structure to find whatever relationships and structure exists within. Critically, the analyst does not have to have any pre-conceived ideas or hypotheses -- they are not specifically looking for any one particular thing. EDA emphasises letting the data inform us and can lead to the discovery of otherwise unexpected relationships, many of which may seem to become completely obvious after discovery. With the knowledge provided by EDA, ideas are generated about what relationships between variables may potentially exist. This then enables the analyst to use these new hypotheses upon which to apply classical inference and rigorously conduct tests with new data. EDA is also related to the field of model diagnostics (MD), where a model can be continuously refined through the visualisation of its fit, its residuals, and the interactions with its variables. Both EDA and MD follow the same framework: Visualise the data, look for patterns that suggest an underlying relationship, and if one is found implement it into the model then continue the exploration of the data. It has long been thought that EDA and statistical inference were worlds apart, but recent work [2, 16] bridges this chasm. Framing a data plot as a test statistic, which when compared to plots of null data, places EDA into the statistical inference recipe. 
The null hypothesis underlying a particular plot, is generically that there is no pattern, and particular types of plots implicitly regulate what “no  pattern” means. For example, a scatter plot of two variables is used to explore for some sort of association, so the implicit hypothesis is that there is no association between the two variables.  The alternative hypothesis is that there is some association, although it is not required to specify precisely the type of association. The plot of the data is placed in a lineup of plots of null data, data generated assuming that the null hypothesis is true. If an impartial observer asked to pick the plot that is different from the rest, picks the data plot this suggests there exists a pattern that is not the result of chance, and the null hypothesis is rejected. For a scatter plot, this departure from the null, might be a single outlier, or few outliers, a non-linear pattern, or clusters, which the human eye detects as more different in this data plot than in any of the nulls. This is the reason why visualisation remains important today, human eyes can detect patterns which would not be detected mathematically.  But eyes need calibration, which the lineup protocol provides.

@@

\begin{figure}[hbtp]
\centering
\includegraphics{cny_line.png}
\caption{Figure 1: Exchange rates for Australian dollars and Chinese yuan against the United States dollar from Mar 30-May 2, 2015, shown as an overlaid line graph. (Both sets of values were standardized to bring them to the same scale for plotting.) Do you think a relationship exists between the two?}
\end{figure}

@@

\begin{figure}[hbtp]
\centering
\begin{subfigure}[t]{0.49\textwidth}
\includegraphics{nzd_scatter.png}
\end{subfigure}
\begin{subfigure}[t]{0.49\textwidth}
\includegraphics{nzd_line.png}
\end{subfigure}
\caption{Figure 2: Exchange rates for Australian and NZ dollars against US dollar, from Mar 30-May 2, 2015, shown using a scatter plot (left) and overlaid line graph (right). Which one can you read the correlation from more accurately? Point-wise linear correlation between the two series, ignoring autodependence, is 0.64.}
\end{figure}

@@

\begin{figure}[hbtp]
\centering
\includegraphics{murders.jpg}
\caption{Figure 3 [19]: Murders in New York State (Orange) and New York City (Blue). The difference between the two represents murders outside of the city. What size would you judge the difference to be in 1994?}
\end{figure}

@@

With the general lack of inference present in exploratory data analysis, how do we attempt to control error in hypothesis testing? In order to use visualisation effectively, we will get the most benefit if the type of graphic display chosen is best suited to the task at hand, be it EDA, MD, or even the presentation of results. We then need to find the statistical power of graphics, the ability of a graphical display to convey information on the structure of the data within. This can be found by using the lineup protocol developed in 2009 by Buja et al. [2], which is easily implemented with the nullabor R package. They created a statistically rigorous framework with properties explored primarily in Majumder et. al [16] and further in Hofmann et. al [11] that allows us to conduct these hypothesis tests with visual inference, thus to some degree allowing the conjoining of EDA with classical inferential statistics.
The lineup protocol is inspired in part by the police lineup [27]. We place the plot of the 'true' data generated with some underlying structure (The criminal) in amongst plots of data that were generated from a null distribution (The innocent people). If the real data plot is selected by an uninvolved observer as being most different from the other plots there is evidence that the structure of that data has led to a significant difference in the plot (That the criminal is sufficiently different from the innocents). This constitutes a rejection of the null hypothesis. If one of the null plots is chosen instead then either the plot design did not have the sufficient power to display the true relationship (A Type 2 error in EDA), or the null plot exhibited a strong relationship that was generated randomly and an analyst that saw this plot and decided that the data had some structure would’ve committed a Type 1 error. p_i, the probability that plot i is chosen in the lineup depends not only on the plot design d, but also the signal strength, q_i, of that plot and of every other competing plot in the lineup. It can be defined as some unknown function f_(i,d) (q_1,…,q_20) [11]. If the true plot is detected, it indicates that the plot design could convey the desired information about the underlying structure and that the true plot has a greater signal strength than those generated under the null distribution. The plot design that has a human observer selecting the true plot the most often will thus minimise both Type 1 error and Type 2 error in EDA hypothesis testing.
However, there is the possibility that the true plot was picked by chance, that we have committed a Type 1 error in the lineup test. For a lineup of m plots, the probability of selecting any plot when they are all generated by the null distribution is 1/m, setting the Type 1 error rate, α, of a lineup hypothesis test to equal 1/m. To control our error rate, it is initially obvious that we can simply change m, with an increase in null lineups having an inverse reduction in m. We recruit human observers to judge the lineups, but this can quickly lead to a large cognitive burden to sort through more and more null plots. A much more powerful option is to have multiple different viewers of each lineup, with K observers; the probability that at a particular plot was picked at least x times under the null hypothesis is binomially distributed [2, 11] with:

$$p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)$$

If all $K$ observers pick the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . This research used $m= 20$, giving us a significance level (and Type 1 error rate) of $α = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.

\begin{figure}[hbtp]
\centering
<<lineup.example, dependson='setup', fig.width=6, fig.height=6, outwidth='0.7//textwidth', echo=FALSE>>=
set.seed(1234)
a <- gen_true_data(24, 0.95)
td <- melt(a, id="t")
b <- gen_null(24)
b$.n <- NULL
nd_l <- melt(b, id="t")
nd_l$.n <- rep(1:19, rep(24, 19))
pos <- 7
td <- data.frame(.n=rep(pos, nrow(td)), td)
lg <- nd_l$.n < pos
nd_l1 <- data.frame(.n=nd_l$.n[lg], t=nd_l$t[lg], variable=nd_l$variable[lg], 
                    value=nd_l$value[lg])
nd_l2 <- data.frame(.n=nd_l$.n[!lg], t=nd_l$t[!lg], variable=nd_l$variable[!lg], 
                    value=nd_l$value[!lg])
nd_l2$.n <- nd_l2$.n + 1
d <- rbind(nd_l1, td, nd_l2)

ggplot(data=d, aes(x=t, y=value, colour=variable)) + geom_line() + facet_wrap(~.n, ncol=4, scales="free_y")  +
    theme_bw() + theme(axis.title.x = element_blank(),
                       axis.title.y = element_blank(),
                       axis.text.x = element_blank(),
                       axis.text.y = element_blank(),
                       axis.ticks = element_blank(),
                       legend.position="none")
@
\caption{Figure 4: A lineup ($m = 20$) of line graphs. Which plot shows the series with the strongest correlation?}
\end{figure}

Essentially, the more time our subjects pick the correct plot out of the lineup, the more confident we are that the real plot was visually significantly different to the nulls; and that the type of graphic involved has the power to display association. The lineup effectively allows us to conduct a hypothesis test on visual features, with the competing hypotheses:
\begin{indent}
H_0 : There is no association between the two series (evidence for H0: the associated data is indistinguishable from the nulls).
H_1: There is an association between the two series (evidence against H0: the associated data can be distinguished from the nulls).
\end{indent}

@@

>>>>

\documentclass[12pt]{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage{wrapfig,float}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{afterpage}

\graphicspath{{figure/}}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcommand{\blue}[1]{{\color{blue} #1}}
\newcommand{\hh}[1]{{\color{magenta} #1}}
\newcommand{\dc}[1]{{\color{orange} #1}}
\newcommand{\green}[1]{{\color{cyan} #1}}

\newtheorem{thm}{Theorem}[section]
\newtheorem{dfn}{Definition}[section]
\newtheorem{cor}{Corollary}[thm]
\newtheorem{con}{Conjecture}[thm]

% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\SweaveOpts{concordance=TRUE}

@@

\if0\blind
{
\author{Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
and\\
Nathaniel Tomasetti\\
Department of Econometrics and Business Statistics, Monash University\\}
  \maketitle
} \fi
\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Comparing the Power of Plot Designs to Reveal Correlation}
\end{center}
  \medskip
} \fi

\bigskip

@@

\newpage
\spacingset{1.45} % DON'T change the spacing!
\newpage
<<setup, echo=FALSE, results=hide, message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)

@@

In order to work with data, it first must be understood. Statistical inference requires hypotheses to be established prior to data collection, but often data is collected first.
This is especially so today, for vast databases that have been assembled in the big data era, that now need the data scientist to unravel the meaning of the numbers. 
Without preset hypotheses to test, the power of statistical inference is impotent, and without hypotheses the data analyst can stumble blindly trying to build up models of structure in the data. 
To understand data requires good visualisation. This idea was formed as early as the 18th century, when William Playfair institutionalized the then revolutionary idea of graphing government and economic data. 
Far easier than reading tables of numbers, these ideas were powerful, and by providing the basic building blocks for plotting statistical data, his graphic designs became a conduit for the communication of otherwise complex information [20]. Since then, advances in computing power have allowed statistical graphics to flourish, spearheaded by Tukey in 1965 into the new domain of exploratory data analysis (EDA) [7, 23]. 
EDA can be thought of as a well understood [3, 25] set of tools and techniques required to visualise information, to physically see what the data contains. In particular, it incorporates a free roaming approach, where the analyst is able to explore structure to find whatever relationships and structure exists within. 
Critically, the analyst does not have to have any pre-conceived ideas or hypotheses -- they are not specifically looking for any one particular thing. EDA emphasises letting the data inform us and can lead to the discovery of otherwise unexpected relationships, many of which may seem to become completely obvious after discovery. 
With the knowledge provided by EDA, ideas are generated about what relationships between variables may potentially exist. This then enables the analyst to use these new hypotheses upon which to apply classical inference and rigorously conduct tests with new data.
EDA is also related to the field of model diagnostics (MD), where a model can be continuously refined through the visualisation of its fit, its residuals, and the interactions with its variables. Both EDA and MD follow the same framework: Visualise the data, look for patterns that suggest an underlying relationship, and if one is found implement it into the model then continue the exploration of the data. 
It has long been thought that EDA and statistical inference were worlds apart, but recent work [2, 16] bridges this chasm. Framing a data plot as a test statistic, which when compared to plots of null data, places EDA into the statistical inference recipe. 
The null hypothesis underlying a particular plot, is generically that there is no pattern, and particular types of plots implicitly regulate what “no  pattern” means. For example, a scatter plot of two variables is used to explore for some sort of association, so the implicit hypothesis is that there is no association between the two variables.
The alternative hypothesis is that there is some association, although it is not required to specify precisely the type of association. 
The plot of the data is placed in a lineup of plots of null data, data generated assuming that the null hypothesis is true. If an impartial observer asked to pick the plot that is different from the rest, picks the data plot this suggests there exists a pattern that is not the result of chance, and the null hypothesis is rejected. 
For a scatter plot, this departure from the null, might be a single outlier, or few outliers, a non-linear pattern, or clusters, which the human eye detects as more different in this data plot than in any of the nulls. This is the reason why visualisation remains important today, human eyes can detect patterns which would not be detected mathematically.  But eyes need calibration, which the lineup protocol provides.

@@

Insert Figure 1

@@

Insert Figure 2

@@

Insert Figure 3

@@

With the general lack of inference present in exploratory data analysis, how do we attempt to control error in hypothesis testing? In order to use visualisation effectively, we will get the most benefit if the type of graphic display chosen is best suited to the task at hand, be it EDA, MD, or even the presentation of results.
We then need to find the statistical power of graphics, the ability of a graphical display to convey information on the structure of the data within. This can be found by using the lineup protocol developed in 2009 by Buja et al. [2], which is easily implemented with the nullabor R package.
They created a statistically rigorous framework with properties explored primarily in Majumder et. al [16] and further in Hofmann et. al [11] that allows us to conduct these hypothesis tests with visual inference, thus to some degree allowing the conjoining of EDA with classical inferential statistics.
The lineup protocol is inspired in part by the police lineup [27]. We place the plot of the 'true' data generated with some underlying structure (The criminal) in amongst plots of data that were generated from a null distribution (The innocent people). If the real data plot is selected by an uninvolved observer as being most different from the other plots there is evidence that the structure of that data has led to a significant difference in the plot (That the criminal is sufficiently different from the innocents). 
This constitutes a rejection of the null hypothesis. If one of the null plots is chosen instead then either the plot design did not have the sufficient power to display the true relationship (A Type 2 error in EDA), or the null plot exhibited a strong relationship that was generated randomly and an analyst that saw this plot and decided that the data had some structure would’ve committed a Type 1 error. 
$p_{i}$, the probability that plot $i$ is chosen in the lineup depends not only on the plot design $d$, but also the signal strength, $q_{i}$, of that plot and of every other competing plot in the lineup. It can be defined as some unknown function $f_{i,d} (q_{1},\dots,q_{20})$ [11]. 
If the true plot is detected, it indicates that the plot design could convey the desired information about the underlying structure and that the true plot has a greater signal strength than those generated under the null distribution. The plot design that has a human observer selecting the true plot the most often will thus minimise both Type 1 error and Type 2 error in EDA hypothesis testing.
However, there is the possibility that the true plot was picked by chance, that we have committed a Type 1 error in the lineup test. For a lineup of $m$ plots, the probability of selecting any plot when they are all generated by the null distribution is $1/m$, setting the Type 1 error rate, $\alpha$, of a lineup hypothesis test to equal $1/m$. 
To control our error rate, it is initially obvious that we can simply change $m$, with an increase in null lineups having an inverse reduction in $\alpha$. We recruit human observers to judge the lineups, but this can quickly lead to a large cognitive burden to sort through more and more null plots. A much more powerful option is to have multiple different viewers of each lineup, with $K$ observers; the probability that at a particular plot was picked at least $x$ times under the null hypothesis is binomially distributed [2, 11] with:

$$p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)$$
If all $K$ observers pick the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . This research used $m= 20$, giving us a significance level (and Type 1 error rate) of $α = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.
\begin{figure}[hbtp]
  \centering
<<lineup.example, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(1234)
a <- gen_true_data(24, 0.95)
td <- melt(a, id="t")
b <- gen_null(24)
b$.n <- NULL
nd_l <- melt(b, id="t")
nd_l$.n <- rep(1:19, rep(24, 19))
pos <- 7
td <- data.frame(.n=rep(pos, nrow(td)), td)
lg <- nd_l$.n < pos
nd_l1 <- data.frame(.n=nd_l$.n[lg], t=nd_l$t[lg], variable=nd_l$variable[lg], 
                    value=nd_l$value[lg])
nd_l2 <- data.frame(.n=nd_l$.n[!lg], t=nd_l$t[!lg], variable=nd_l$variable[!lg], 
                    value=nd_l$value[!lg])
nd_l2$.n <- nd_l2$.n + 1
d <- rbind(nd_l1, td, nd_l2)

plot <- ggplot(data=d, aes(x=t, y=value, colour=variable)) + geom_line() + facet_wrap(~.n, ncol=4, scales="free_y")  +
    theme_bw() + theme(axis.title.x = element_blank(),
                       axis.title.y = element_blank(),
                       axis.text.x = element_blank(),
                       axis.text.y = element_blank(),
                       axis.ticks = element_blank(),
                       legend.position="none")
print(plot)
@
  \caption{A lineup ($m = 20$) of line graphs. Which plot shows the series with the strongest correlation?}
  \label{lineup:example}
\end{figure}

Essentially, the more time our subjects pick the correct plot out of the lineup, the more confident we are that the real plot was visually significantly different to the nulls; and that the type of graphic involved has the power to display association. The lineup effectively allows us to conduct a hypothesis test on visual features, with the competing hypotheses:

\begin{itemize}
  \item $H_{0}$ : There is no association between the two series (evidence for $H_{0}$: the associated data is indistinguishable from the nulls).
  \item $H_{1}$: There is an association between the two series (evidence against $H_{0}$: the associated data can be distinguished from the nulls).
\end{itemize}

@@

To allow for a greater control of distributions of the data in the lineups, we use simulated data. The data must contain autocorrelation for there to be meaningful information in the time dimension. To fulfil this requirement, each plot has a standardised pair of AR(1) models defined by:

$$Y_{1,t} = \beta \times Y_{1,t-1} + e_{1,t}$$

$$Y_{2,t} = \beta \times Y_{2,t-1} + e_{2,t}$$

With $t \in \left\{12, 24, 48, 96\right\}$ and $e_{t}$ being generated with from a bivariate normal distribution:

$$~N(\mu= \left| \begin{array}{c}0 \\0 \end{array} \right|,\hspace{4mm}\Sigma = \left| \begin{array}{cc}1 & \rho \\ \rho & 1\end{array} \right|)$$

The variation in the covariance parameter  allows for control in the correlation in the pairs of data. The null data has no structure, so uncorrelated pairs are produced by setting $\rho = 0$. True data pairs are generated with $\rho = \pm 0.3, 0.5, 0.7, 0.9$. Pilot testing found that the rate of picking the true data plot fell to approximately $5\%$, the $1/m$ rate of randomly picking the true plot, for pairs with correlation below $\rho = 0.3$. Further to this, relationships weaker than this point are often of little interest to analysts. Hence we found it unnecessary to test a more full range of correlations. Simulating data does not produce correlations exactly equal to $\rho$, so imulated data pairs were accepted as real data if pairwise correlation between the two series is within 0.015 of the desired value of $\rho$. The null pairs often had correlation generated spuriously through the simulation process. This was particularly problematic for generating small sample data, but setting, to a maximum of 0.5, for both true and null data reduced spurious correlation to manageable levels.
For particularly large time series, the AR(1) model can create particularly 'jagged' lines that may be unrealistic for many actual applications. To counter this, additional sets of data were created with a Hodrick-Prescott filter to capture the trend of the model and remove the noise in the cycle component. This filter creates a trend by penalising both errors in smoothness and in fit. The trend of the series, $\tau_{t}$ is defined by the equation:

$$\min_{\tau}(\sum_{t=1}^{T}(y_{t}-\tau_{t})^2 + \lambda \sum_{t=2}^{T-1}[(\tau_{t+1}-\tau_{t})-(\tau_{t}-\tau_{t-1})]^2)$$

Setting $\lambda = 1$ produced a smoothing that is more consistent with real world time series applications and did not introduce excessive spurious null correlation. However, for $t = 12, 24$, any form of smoothing was unfeasible without introducing extreme null correlation, hence smoothing was only used for $t = 48, 96$. Each generated lineup was produced as both a scatter plot and an overlaid line graph.

\begin{figure}[hbtp]
  \centering
<<smoothing, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(123)
uns <- gen_true_data(96, 0.8)
uns_l <- melt(uns, id="t")
p1a <- ggplot(data=uns_l, aes(x=t, y=value, colour=variable)) + geom_line() +labs(title="Unsmoothed Overlaid Lines") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p3a <- ggplot(data=uns, aes(x=X1, y=X2)) + geom_point()+ labs(title="Unsmoothed Scatterplot)") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")

set.seed(123)
sm <- gen_true_data(96, 0.8, smoothed = TRUE)
sm_l <- melt(sm, id="t")
p2a <- ggplot(data=sm_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Smoothed Overlaid Lines") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p4a <- ggplot(data=sm, aes(x=X1, y=X2)) + geom_point() + labs(title="Smoothed Scatterplot") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")

grid.arrange(p1a, p2a, p3a, p4a, ncol=2)
@
  \caption{The same AR(1) data both smoothed and unsmoothed. t = 96, correlation = 0.86}
  \label{smoothing}
\end{figure}

In this research, we had three basic factors:

\begin{itemize}
  \item Factor 1: Plot design, levels = scatter plot, overlaid time series
  \item Factor 2: Sample size, levels = 12, 24, 48, 96
  \item Factor 3: True plot correlation, levels 
\end{itemize}

Additionally, there was a fourth factor for  = 48, 96:

\begin{itemize}
  \item Factor 4: Smoothed, levels = yes, no
\end{itemize}

Each combination of factors was replicated in three different lineups, giving a total of 288 lineups (2 plots x 8 correlations x 6 sample sizes & smoothness combinations). 
The position of the actual data plot in the lineup was randomized. The order that the factors are presented to a subject were also randomized, and each subject saw 5 scatter plots and 5 overlaid lines at a range of all absolute correlation levels. Each subject did not see the same data more than once. Subjects were asked to pick the plot in the lineup that has the strongest association, with basic examples of negatively and positively correlated plots. They also answered with reasons for choosing the plot, and the confidence that they have that this really is the plot showing the strongest correlation. Each subject evaluated 10 lineups, plus a further two trials to eliminate people attempted to game the system. If a trial was not answered correctly, the subject could not evaluate the remaining lineups. If it was answered correctly, the trial data was discarded and the subject moved onto the rest of the lineups. Earlier work [15] has found that repeated evaluations from a subject does not increase their ability to detect the true plot from a lineup, justifying our treatment of the results as independent evaluations. We recruited subjects from Amazon's Mechanical Turk service and ensured that each lineup was viewed by multiple subjects. In total there were 2088 lineup evaluations, however to better control each subject’s individual visual ability we removed data from subjects with less than ten individual evaluations leaving us with 1684 evaluated lineups for the analysis.

\begin{figure}[hbtp]
  \centering
<<designs, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE>>=
set.seed(101)
a <- gen_true_data(24, 0.95)
b <- gen_true_data(24, -0.95)
ca <- cor(a$X1, a$X2)
cb <- cor(b$X1, b$X2)
a_l <- melt(a, id="t")
b_l <- melt(b, id="t")
p1 <- ggplot(data=a_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Overlaid line graph") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p2 <- ggplot(data=b_l, aes(x=t, y=value, colour=variable)) + geom_line() + labs(title="Overlaid line graph") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p3 <- ggplot(data=a, aes(x=X1, y=X2)) + geom_point() + labs(title="Scatterplot") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
p4 <- ggplot(data=b, aes(x=X1, y=X2)) + geom_point() + labs(title="Scatterplot") +
      theme(axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          legend.position="none")
grid.arrange(p1, p2, p3, p4, ncol=2)
@
  \label{designs}
  \caption{The same two data sets in both plot designs. t=24, correlation (left) = 0.946, correlation (right) = -0.943}
\end{figure}

The web site that the Turk workers accessed to complete the task can be seen at http://104.236.245.153:8080/mahbub/turk18/index.html.  You can also try out the tasks, but your results will not be recorded. (The experiment cost $210 to run.)
